Entropy-SGD: Biasing Gradient Descent Into Wide Valleys

### 2017-01-11 doc/chaudhari.choromanska.ea.iclr17.tex
HS97 para

Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the width of local minima and training loss and show using the Gibbs formalism~\citep{haussler1997mutual} \fix{(their analysis necessitates a few unrealistic assumptions, see their Sec. A1)}{we might consider deleting this} that this leads to improved generalization. However, the effect of hyper-parameters for the constraints they use is intricately tied together and they are difficult to choose even for small problems, \textcolor{red}{as discussed by the authors.} Their formulation finally depends on Hessian-vector products. Driven by the same goal of encouraging the optimizer to favor flat regions in the opimization landscape, we however formulate a different objective function based on local entropy and obtain a more scalable algorithm which does not rely on extraneous and difficult to control hyper-parameters. \fix{Despite fundamental differences, both approaches... }{one sentence about the similarities} Also recent work by~\citet{keskar2016large} is relevant in the context of the main theme of our paper. The authors estimate the loss in a neighborhood of the weights to argue that small batch size in SGD (i.e., larger gradient noise) generalizes better than large mini-batches and also results in significantly flatter minima.

@@

In order to test the efficacy of $\entropysgd$ on recurrent architectures, we train an LSTM network on the Penn Tree Bank (PTB) dataset for word-level text prediction. This dataset contains about one million words divided into a training set of about $930,000$ words, a validation set of $74,000$ words and $82,000$ words with a vocabulary of size $10,000$. As a baseline, we follow the experiments in~\citet{zaremba2014recurrent}. Our network called $\ptblstm$ consists of two layers with $1500$ hidden units, each unrolled for $35$ time steps and a dropout of probability $0.65$; note that this is a large network with about $66$ million weights. It is trained with SGD (without momentum, like the original authors) for $55$ epochs with a learning rate $\eta = 1$ for the first $14$ epochs that is annealed by a factor of $1.15$ after every epoch. \pc{We obtained a word perplexity of $81.125$ on the validation set and $78.232$}{std-dev} on the test set with this setup, which closely matches the results of the original authors.

We run $\entropysgd$ on $\ptblstm$ for $5$ epochs with $L = 5$; note that this results in only $25$ effective epochs. We do not use scoping for this network and instead fix $\g = 10^{-3}$. The initial learning rate of the outer loop is $\eta = 1$ which reduces by a factor of $10$ at each epoch after the third epoch. The SGLD learning rate is fixed to $\eta' = 1$ with $\e = 10^{-4}$. We obtain a word perplexity of $80.116 \pm 0.069$ on the validation set and $77.656 \pm 0.171$ on the test set. As Fig.~\ref{fig:ptb_valid} shows, $\entropysgd$ trains significantly faster than SGD and also achieves a slightly better generalization perplexity.

>>>>

Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the training loss and the width of local minima and show using the Gibbs formalism~\citep{haussler1997mutual} that this leads to improved generalization. As the authors discuss, the effect of hyper-parameters for the constraints they use is intricately tied together and they are difficult to choose even for small problems. Our local entropy based objective naturally balances the energetic term (training loss) and the entropic term (width of the valley) and also results in a more scalable algorithm. The scope $\g$ affords more control on the optimization process than hard constraints and as we discuss in Sec.~\ref{ss:scoping}, this provides significant computational advantages. Let us note that the Gibbs formalism introduced in~\citet{hochreiter1997flat} seems to be a very promising direction to understanding generalization in deep networks, it bears supercial similarities to the uniform stability analysis~\citep{haussler1997mutual} we employ. Recent work by~\citet{keskar2016large} is relevant in the context of the main theme of our paper. The authors estimate the loss in a neighborhood of the weights to argue that small batch size in SGD (i.e., larger gradient noise) generalizes better than large mini-batches and also results in significantly flatter minima.

@@

In order to test the efficacy of $\entropysgd$ on recurrent architectures, we train an LSTM network on the Penn Tree Bank (PTB) dataset for word-level text prediction. This dataset contains about one million words divided into a training set of about $930,000$ words, a validation set of $74,000$ words and $82,000$ words with a vocabulary of size $10,000$. As a baseline, we follow the experiments in~\citet{zaremba2014recurrent}. Our network called $\ptblstm$ consists of two layers with $1500$ hidden units, each unrolled for $35$ time steps and a dropout of probability $0.65$; note that this is a large network with about $66$ million weights. It is trained with SGD (without momentum, like the original authors) for $55$ epochs with a learning rate $\eta = 1$ for the first $14$ epochs that is annealed by a factor of $1.15$ after every epoch. We obtained a word perplexity of $81.43 \pm 0.2$ on the validation set and $78.6 \pm 0.26$ on the test set with this setup, which closely matches the results of the original authors.

We run $\entropysgd$ on $\ptblstm$ for $5$ epochs with $L = 5$, note that this results in only $25$ effective epochs. We do not use scoping for this network and instead fix $\g = 10^{-3}$. The initial learning rate of the outer loop is $\eta = 1$ which reduces by a factor of $10$ at each epoch after the third epoch. The SGLD learning rate is fixed to $\eta' = 1$ with $\e = 10^{-4}$. We obtain a word perplexity of $80.116 \pm 0.069$ on the validation set and $77.656 \pm 0.171$ on the test set. As Fig.~\ref{fig:ptb_valid} shows, $\entropysgd$ trains significantly faster than SGD and also achieves a slightly better generalization perplexity.
